Controlled Experiments for Word Embeddings

### 2015-10-09 article/main.tex
add reference to wsd pseudoword

\begin{abstract}
	An experimental approach to studying the properties of word
        embeddings is proposed.  Controlled experiments, achieved
        through modifications of the training corpus, permit the
        demonstration of direct relations between word properties and
        word vector direction and length.  The approach is demonstrated
        using the word2vec CBOW model with experiments that
        independently vary word frequency and word co-occurrence noise.
        The experiments reveal that word vector length depends more or
        less linearly on both word frequency and the level of noise in
        the co-occurrence distribution of the word.  The coefficients of
        linearity depend upon the word.  The special vector in feature
        space, defined by the (artificial) word with pure noise in its
        co-occurrence distribution, is found to be small but non-zero.
\end{abstract}

@@

In this paper, we wish to study in detail to what extent these two
factors determine word vectors.  For a given corpus, both term
frequency and co-occurrence are, of course, fixed and it is not obvious
how to unravel these dependencies in an unambiguous, objective manner.
In particular, it is difficult to establish the distinctiveness of the
contexts in which a word is used.  To overcome these problems, we
propose to modify the training corpus in a controlled fashion.  To
this end, we insert new pseudowords into the corpus with varying frequencies
and varying levels of noise in their co-occurrence distributions.  By
modeling the frequency and co-occurrence distributions of these pseudowords
on existing words in the corpus, we are able to study their effect on
word vectors independently of one another.  We can thus study a family
of pseudowords that all appear in the same context, but with different
frequencies, or study a family of pseudowords that all have the same
frequency, but appear in a different number of contexts.  Starting from
the limited number of contexts in which a word appears in the original
corpus, we can increase this number by interspersing the word in arbitrary
contexts at random.  The word thus looses its significance in a
controlled way.  Although we present our approach using the word2vec
CBOW model, these and related experiments could equally well be carried
out for other word embedding methods such as the word2vec skip-gram
model \cite{DistRepns,EfficientEstimation}, GloVe
\cite{pennington2014glove}, and SENNA \cite{collobert-2011}.

We show that the length of the word vectors generated by the CBOW model
depends more or less linearly on both word frequency and level of noise
in the co-occurrence distribution of the word.  In both cases, the
coefficient of linearity depends upon the word.  If the co-occurrence
distribution is fixed, then word vector length increases with word
frequency.  If, on the other hand, word frequency is held constant, then
word vector length decreases as the level of noise in the co-occurrence
distribution of the word is increased.  We show furthermore that the
direction of the word vectors is independent of word frequency and only
moderately depends upon co-occurrence noise.  When noise is added to the
co-occurrence distribution of a word, the corresponding vector smoothly
interpolates between the original word vector and a small vector
perpendicular to it that represents a word with pure noise in its
co-occurrence distribution.  Surprisingly, this special vector in the feature
space, obtained by interspersing a pseudoword uniformly at random throughout
the corpus with a frequency sufficiently large to sample all contexts, is
non-zero.

@@

\begin{section}{Related work}\label{related-work}
In recent work on the evaluation of word embeddings, Schnabel, Labutov, Mimno
and Joachims \cite{schnabelemnlp2015} trained logistic regression models to
predict whether a word was rare or frequent given only the direction of its
word vector.  The prediction accuracy was then measured as a function of word
embedding method and of the threshold for word rarity.  It was found in the
case of word2vec CBOW that word vector direction could be used to distinguish
very rare words from all other words. The authors claimed further that word
vector direction must encode frequency directly, and not indirectly via
semantic information.  Our experimental finding that word vector direction is
independent of word frequency contradicts this claim in the case of word2vec
CBOW.

@@

\begin{figure}
	\begin{mdframed}
	\input{word-frequency-experiment-text-cat.tex}
	\end{mdframed}
	\caption{Example sentences modified in the word frequency
          experiment as per Section~\ref{WFVEexisting}, where the word
          \word{cat} is replaced with pseudowords \word{CAT\_i} using the
          truncated geometric distribution (\ref{distro}) with
          $p=\tfrac{1}{2}$ and $n=20$.}
	\label{fig:word-frequency-experiment-text-cat}
\end{figure}

@@

\begin{subsection}{Tokens derived from an artificial, meaningless word}
\label{WFVEmeaningless}
Whereas the pseudowords introduced above all replace an existing word that
carries a meaning, we now include for comparison a high-frequency,
meaningless word.  We choose to introduce an artificial, entirely
meaningless word \word{VOID} into the corpus, rather than choose an
existing (stop)word whose meaninglessness is only supposed.  To achieve
this, we intersperse the word uniformly at random throughout the corpus
so that its relative frequency is $0.005$.  The co-occurrence
distribution of \word{VOID} thus coincides with the unconditional word
distribution.  The replacement procedure is then performed for this
word, using the same values for $p$ and $n$ as above.
Figure~\ref{fig:word-frequency-experiment-text-void} shows the effect of
these modifications on a sample text, where a higher relative frequency
of $0.05$ is used instead for illustrative purposes.
%Although the probability of inserting the pseudoword
%\word{VOID\_1} is twice that of inserting the pseudoword \word{VOID\_2} and four
%times that of inserting the pseudoword \word{VOID\_3}, these three pseudowords all
%appear once in this small piece of text.

@@

\begin{figure}
	\begin{mdframed}
	\input{cooccurrence-noise-experiment.tex}
	\end{mdframed}
	\caption{Example sentences modified for the co-occurrence noise
          experiment, where the word \word{cat} was chosen for
          replacement.  The pseudowords were generated using the distribution
          (\ref{distro2}) with $n=7$.}
	\label{fig:co-occurrence-noise-experiment-text}
\end{figure}

\begin{subsection}{Experimental results}
%
Figure~\ref{fig:co-occurrence-noise-heatmap} shows the cosine similarity
of pairs of vectors representing some of the pseudowords used in this
experiment.  Remember that the first pseudoword ($i=1$) in a family is
without noise in its co-occurrence distribution, while the last one
($i=n$, with $n=7$) stands for pure noise and has therefore no relation
anymore with the word it derives from.  The figure demonstrates that the
vectors within a family only moderately deviate from the original
direction defined by the first pseudoword ($i=1$) when noise is added to the
co-occurrence distribution.  For $1<i<7$, the deviation typically
increases with the proportion of noise.  The vector of the last pseudoword
($i=n$), associated with pure noise, is seen within each of the families
to point in a completely different direction, more or less perpendicular
to the original one.  To understand this interpolating behavior, recall
from Section~\ref{WFVE-results} that the vector for the entirely
meaningless word \word{VOID} is small but non-zero.  Since the noise
distribution coincides with the co-occurrence distribution of
\word{VOID}, the vectors for the experiment words must tend to the word
vector for \word{VOID} as the proportion of noise in their co-occurrence
distributions approaches $1$.
This convergence to a common point is only indistinctly apparent in
Figure~\ref{fig:co-occurrence-noise-heatmap} as the frequency of the experiment
pseudowords is insufficient to sample the full variety of the contexts of
\word{VOID}, i.e.  all contexts (see Section \ref{WFVE-direction}).

@@

The co-occurrence distribution of \word{VOID} is the unconditional
frequency distribution, and in this sense pure background noise.  Thus
the word vector of \word{VOID} is a special point in the feature space.
Figure~\ref{fig:word-frequency-experiment-graph} shows that this point
is not at the origin of the feature space, i.e., is not the zero vector.
The origin, however, is implicitly the point of reference in word2vec
word similarity tasks.  This raises the question of whether improved
performance on similarity tasks could be achieved by transforming the
feature space or modifying the model such that the representation of
pure noise, i.e., the vector for \word{VOID}, is at the origin of the
transformed feature space.

%\textcolor{red}{can it be argued that the feature space must be fundamentally Euclidean, given that the path marked by the vectorisations of intermediate words is a straight line? i.e., if such paths maybe considered as ``geodesics'' between words, and we show that these geodesics appear as straight lines when an Euclidean metric is imposed, then suggests that the Euclidean metric is in fact a good choice.  what would happen if we chose another metric?}

>>>>

\begin{abstract}
	An experimental approach to studying the properties of word
        embeddings is proposed.  Controlled experiments, achieved
        through modifications of the training corpus, permit the
        demonstration of direct relations between word properties and
        word vector direction and length.  The approach is demonstrated
        using the word2vec CBOW model with experiments that
        independently vary word frequency and word co-occurrence noise.
        The experiments reveal that word vector length depends more or
        less linearly on both word frequency and the level of noise in
        the co-occurrence distribution of the word.  The coefficients of
        linearity depend upon the word.  The special point in feature
        space, defined by the (artificial) word with pure noise in its
        co-occurrence distribution, is found to be small but non-zero.
\end{abstract}

@@

In this paper, we wish to study in detail to what extent these two
factors determine word vectors.  For a given corpus, both term
frequency and co-occurrence are, of course, fixed and it is not obvious
how to unravel these dependencies in an unambiguous, objective manner.
In particular, it is difficult to establish the distinctiveness of the
contexts in which a word is used.  To overcome these problems, we
propose to modify the training corpus in a controlled fashion.  To
this end, we insert new tokens into the corpus with varying frequencies and
varying levels of noise in their co-occurrence distributions.  By modeling the
frequency and co-occurrence distributions of these tokens, or 
\textit{pseudowords}\footnote{We refer to these tokens as
pseudowords, since their properties are modeled upon words in the
lexicon and because our corpus modification approach is reminiscent of the
pseudoword approach for generating labeled data for word sense disambiguation
tasks in \cite{gale1992work}.}, on existing words in the corpus, we are able to
study their effect on word vectors independently of one another.  We can thus
study a family of pseudowords that all appear in the same context, but with
different frequencies, or study a family of pseudowords that all have the same
frequency, but appear in a different number of contexts.  Starting from the
limited number of contexts in which a word appears in the original corpus, we
can increase this number by interspersing the word in arbitrary contexts at
random.  The word thus looses its significance in a controlled way.  Although
we present our approach using the word2vec CBOW model, these and related
experiments could equally well be carried out for other word embedding methods
such as the word2vec skip-gram model \cite{DistRepns,EfficientEstimation},
GloVe \cite{pennington2014glove}, and SENNA \cite{collobert-2011}.

We show that the length of the word vectors generated by the CBOW model
depends more or less linearly on both word frequency and level of noise
in the co-occurrence distribution of the word.  In both cases, the
coefficient of linearity depends upon the word.  If the co-occurrence
distribution is fixed, then word vector length increases with word
frequency.  If, on the other hand, word frequency is held constant, then
word vector length decreases as the level of noise in the co-occurrence
distribution of the word is increased.  We show furthermore that the
direction of the word vectors is independent of word frequency and only
moderately depends upon co-occurrence noise.  When noise is added to the
co-occurrence distribution of a word, the corresponding vector smoothly
interpolates between the original word vector and a small vector
perpendicular to it that represents a word with pure noise in its
co-occurrence distribution.  Surprisingly, the special point in feature
space, obtained by interspersing a pseudoword uniformly at random
throughout the corpus with a frequency sufficiently large to sample all
contexts, is non-zero.

@@

\begin{section}{Related work}\label{related-work}
In recent work on the evaluation of word embeddings, Schnabel et
al.~\cite{schnabelemnlp2015} trained logistic regression models to
predict whether a word was rare or frequent given only the direction of
its word vector.  For various word embedding methods, the prediction
accuracy was measured as a function of the threshold for word rarity.
It was found in the case of word2vec CBOW that word vector direction
could be used to distinguish very rare words from all other words. The
authors claimed further that word vector direction must encode frequency
directly, and not indirectly via semantic information.  Our experimental
finding that word vector direction is independent of word frequency
contradicts this claim in the case of word2vec CBOW.

@@

\begin{figure}
	\begin{mdframed}
	\input{word-frequency-experiment-text-cat.tex}
	\end{mdframed}
	\caption{Example sentences modified in the word frequency
          experiment as per Section~\ref{WFVEexisting}, where the word
          \word{cat} is replaced with pseudowords \word{CAT\_i} using the
          truncated geometric distribution~(\ref{distro}) with
          $p=\tfrac{1}{2}$ and $n=20$.}
	\label{fig:word-frequency-experiment-text-cat}
\end{figure}

@@

\begin{subsection}{Pseudowords derived from an artificial, meaningless word}
\label{WFVEmeaningless}
Whereas the pseudowords introduced above all replace an existing word that
carries a meaning, we now include for comparison a high-frequency,
meaningless word.  We choose to introduce an artificial, entirely
meaningless word \word{VOID} into the corpus, rather than choose an
existing (stop)word whose meaninglessness is only supposed.  To achieve
this, we intersperse the word uniformly at random throughout the corpus
so that its relative frequency is $0.005$.  The co-occurrence
distribution of \word{VOID} thus coincides with the unconditional word
distribution.  The replacement procedure is then performed for this
word, using the same values for $p$ and $n$ as above.
Figure~\ref{fig:word-frequency-experiment-text-void} shows the effect of
these modifications on a sample text, where a higher relative frequency
of $0.05$ is used instead for illustrative purposes.
%Although the probability of inserting the pseudoword
%\word{VOID\_1} is twice that of inserting the pseudoword \word{VOID\_2} and four
%times that of inserting the pseudoword \word{VOID\_3}, these three pseudowords all
%appear once in this small piece of text.

@@

\begin{figure}
	\begin{mdframed}
	\input{cooccurrence-noise-experiment.tex}
	\end{mdframed}
	\caption{Example sentences modified for the co-occurrence noise
          experiment, where the word \word{cat} was chosen for
          replacement.  The pseudowords were generated using the
          distribution~(\ref{distro2}) with $n=7$.}
	\label{fig:co-occurrence-noise-experiment-text}
\end{figure}

\begin{subsection}{Experimental results}
%
Figure~\ref{fig:co-occurrence-noise-heatmap} shows the cosine similarity
of pairs of vectors representing some of the pseudowords used in this
experiment.  Remember that the first pseudoword~($i=1$) in a family is
without noise in its co-occurrence distribution, while the last one
($i=n$, with $n=7$) stands for pure noise and has therefore no relation
anymore with the word it derives from.  The figure demonstrates that the
vectors within a family only moderately deviate from the original
direction defined by the first pseudoword ($i=1$) when noise is added to the
co-occurrence distribution.  For $1<i<7$, the deviation typically
increases with the proportion of noise.  The vector of the last pseudoword
($i=n$), associated with pure noise, is seen within each of the families
to point in a completely different direction, more or less perpendicular
to the original one.  To understand this interpolating behavior, recall
from Section~\ref{WFVE-results} that the vector for the entirely
meaningless word \word{VOID} is small but non-zero.  Since the noise
distribution coincides with the co-occurrence distribution of
\word{VOID}, the vectors for the experiment words must tend to the word
vector for \word{VOID} as the proportion of noise in their co-occurrence
distributions approaches $1$.
This convergence to a common point is only indistinctly apparent in
Figure~\ref{fig:co-occurrence-noise-heatmap}, as the frequency of the experiment
pseudowords is insufficient to sample the full variety of the contexts of
\word{VOID}, i.e., all contexts (see Section~\ref{WFVE-direction}).

@@

The co-occurrence distribution of \word{VOID} is the unconditional
frequency distribution, and in this sense pure background noise.  Thus
the word vector of \word{VOID} is a special point in the feature space.
Figure~\ref{fig:word-frequency-experiment-graph} shows that this point
is not at the origin of the feature space, i.e., is not the zero vector.
The origin, however, is implicitly the point of reference in word2vec
word similarity tasks.  This raises the question of whether improved
performance on similarity tasks could be achieved by transforming the
feature space or modifying the model such that the representation of
pure noise, i.e., the vector for \word{VOID}, is at the origin of the
transformed feature space.
\end{section}
